---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gluedown

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/gluedown)](https://CRAN.R-project.org/package=gluedown)
<!-- badges: end -->

The `gluedown` package helps wrap character vectors in markdown syntax.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kiernann/gluedown")
```

## Usage

By default, many `gluedown` functions simply take a character vector and return that same vector
with each element wrapped in the corresponding markdown syntax. With the `cat` argument, these
wrapped vectors can by easily concatenated, separated with a newline, and printed; this is useful
in combination with the `results='asis'` RMarkdown chunk option to directly print markdown blocks.

```{r prepare, warning=FALSE, message=FALSE}
library(gluedown)
library(dplyr)
library(rvest)
states <- sample(state.name, 3)
chores <- c("Wake up", "Eat breakfast", "Brush teeth")
preamble <- html_text(html_nodes(read_html("https://plaintext.us/constitution"), "p"))[[1]]
```

```{r example_lists, results='asis'}
md_bullet(states, cat = TRUE)
md_list(states, cat = TRUE)
md_task(chores, check = c(1, 3), cat = TRUE)
```

```{r example_quote, results='asis'}
md_quote(preamble, cat = TRUE)
```

```{r example_table}
md_table(band_members)
```
