---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# gluedown

<!-- badges: start -->
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/gluedown)](https://CRAN.R-project.org/package=gluedown)
<!-- badges: end -->

The `gluedown` package helps wrap character vectors in markdown syntax.

## Installation

You can install the development version from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("kiernann/gluedown")
```

## Example

By default, many `gluedown` functions simply create character strings from
character strings. These new vectors can then also be concatenated and printed;
this is useful in combination with the `results='asis'` RMarkdown chunk option.

```{r prepare, warning=FALSE, message=FALSE}
library(gluedown)
library(rvest)
states <- sample(state.name, 3)
chores <- c("Wake up", "Eat Breakfast", "Brush Teeth")
url <- "https://plaintext.us/constitution"
preamble <- html_text(html_nodes(read_html(url), "p"))[[1]]
```

```{r example_lists, results='asis'}
md_bullet(states, cat = TRUE)
md_list(states, cat = TRUE)
md_task(chores, cat = TRUE, checks = c(1, 3))
```

```{r example_quote, results='asis'}
md_quote(preamble, cat = TRUE)
```

The functions are also semi-useful when used inline in Rmarkdown documents.
They can be used to `r md_bold("bold")`, `r md_italic("italicize")`, 
`r md_strike("strike through")`, etc.

## To Do

* Implement the table functionality from `knitr::kable()`
* Improve the vectorization of `md_fence()`
