---
title: "GitHub Flavored Markdown"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
vignette: >
  %\VignetteIndexEntry{GitHub Flavored Markdown}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, warning=FALSE, message=FALSE}
library(gluedown)
library(stringr)
library(rvest)
```

In this vignette, we will explore how the functions in `gluedown` enable users
to transition from R vectors to the kind of formatted markdown text used in
the GitHub Flavored Markdown (GFM) specification. Functions like `md_bullet()`
and `md_quote()` may be used more often, but there are functions for practically
every section of the GFM spec including, some of which have limited practical
value (e.g., `md_blank()` and `md_text()`). 

In each section of this vignette, the GFM spec is quoted first and followed by
`gluedown` function usage. For each R code block containing `gluedown` usage,
the raw text is shows as `#>` comments followed by the rendered HTML version,
which is printed using the `results='asis'` option of the `knitr` package. (To 
achieve this output, each block is actually repeated twice, with the code of the
second result printing block hidden via `echo=FALSE`.)

## 3 Blocks and Inlines

> We can think of a document as a sequence of blocks—structural elements like
paragraphs, block quotations, lists, headings, rules, and code blocks. Some
blocks (like block quotes and list items) contain other blocks; others (like
headings and paragraphs) contain inline content—text, links, emphasized text,
images, code spans, and so on.

> We can divide blocks into two types: container blocks, which can contain other
blocks, and leaf blocks, which cannot.

## 4 Leaf blocks

There are nine leaf block functions:

1. `md_rule()`
1. `md_heading()`
1. `md_setext()`
1. `md_indent()`
1. `md_fence()`
1. `md_convert()`
1. `md_reference()`
1. `md_paragraph()`
1. `md_blank()`
1. `md_table()`

### 4.1 Thematic breaks

> A line consisting of 0-3 spaces of indentation, followed by a sequence of
three or more matching `-`, `_`, or `*` characters, each followed optionally by
any number of spaces or tabs, forms a thematic break.

We can create thematic breaks with the `md_rule()` function, which allows users
to define the type and number of valid characters to use. All these options
create the same `<hr>` HTML tag, but look different in the underlying `.md`
document.

```{r}
md_rule()
```

```{r echo=FALSE, results='asis'}
md_rule()
```

```{r}
md_rule(char = "-", n = 10, space = TRUE)
```

```{r echo=FALSE, results='asis'}
md_rule(char = "-", n = 10, space = TRUE)
```

```{r echo=FALSE, results='asis'}
md_heading("4.2 ATX Headings", level = 3)
```

> An ATX heading consists of a string of characters, parsed as inline content,
between an opening sequence of 1–6 unescaped `#` characters and an optional
closing sequence of any number of unescaped `#` characters... The raw contents
of the heading are stripped of leading and trailing spaces before being parsed
as inline content. The heading level is equal to the number of `#` characters in
the opening sequence.

The heading for this section was created with `md_heading()`.

```{r}
md_heading("4.2 ATX Headings", level = 3)
```

### 4.3 Setext headings

> A setext heading consists of one or more lines of text, each containing at
least one non-whitespace character, with no more than 3 spaces indentation,
followed by a setext heading underline... A setext heading underline is a
sequence of `=` characters or a sequence of `-` characters, with no more than 3
spaces indentation and any number of trailing spaces... The heading is a level 1
heading if `=` characters are used in the setext heading underline, and a level
2 heading if `-` characters are used.

I'm going to forgo printing these headings, just to preserve the structure of this document. The
`md_setext()` function can only be used to create level 1 (`<h1>`) and level 2 (`<h2>`) headings.
A level 1 Setext heading is sometimes used as the very first, initial title heading.

```{r}
md_setext("Headings", level = 1)
```

```{r}
md_setext("4.3 Setext headings", level = 2)
```

```{r}
md_setext("4.3 Setext headings", width = 50)
```

### 4.4 Indented code blocks

> An indented code block is composed of one or more indented chunks separated by
blank lines. An indented chunk is a sequence of non-blank lines, each indented
four or more spaces. The contents of the code block are the literal contents of
the lines, including trailing line endings, minus four spaces of indentation...

This function, along with the fenced code block alternative, are useful for
creating displayed code blocks (as opposed to executed chunks) containing R 
source code, command line content, or any other pre-formatted fixed-width
content.

```{r}
rescale01 <- function(x) {
  rng <- range(x, na.rm = TRUE)
  (x - rng[1]) / (rng[2] - rng[1])
}

# use deparse() to turn unevaluated expressions into character strings.
source <- deparse(rescale01)
# this new object is a regular character vector
print(source)
```

```{r}
md_indent(source)
```

```{r echo=FALSE, results='asis'}
md_indent(source)
```

### 4.5 Fenced code blocks

> A code fence is a sequence of at least three consecutive backtick characters
... or tildes (`~`). (Tildes and backticks cannot be mixed.) A fenced code
block begins with a code fence, indented no more than three spaces...
>
> The line with the opening code fence may optionally contain some text
following the code fence; this is trimmed of leading and trailing whitespace and
called the info string...
>
> The content of the code block consists of all subsequent lines, until a
closing code fence of the same type as the code block began with (backticks or
tildes), and with at least as many backticks or tildes as the opening code
fence....
>
> The content of a code fence is treated as literal text, not parsed as inlines.
The first word of the info string is typically used to specify the language of
the code sample, and rendered in the class attribute of the code tag. However,
this spec does not mandate any particular treatment of the info string.

The man feature of `md_fence()` is the ability to specify the info string which defines the code
language used to highlight syntax. Compare the following output to the one produced by 
`md_indent()`.

```{r}
md_fence(source)
```

```{r echo=FALSE, results='asis'}
md_fence(source)
```

```{r}
md_fence(source, char = "~")
```

```{r echo=FALSE, results='asis'}
md_fence(source, char = "~")
```

```{r}
md_fence("$ sudo apt install r-base-dev", info = "bash")
```

```{r echo=FALSE, results='asis'}
md_fence("$ sudo apt install r-base-dev", info = "bash")
```

### 4.6 HTML blocks

> An HTML block is a group of lines that is treated as raw HTML (and will not be
escaped in HTML output).

Markdown rendering engines (like the one used to convert this `.Rmd` document
to `.md` and then to `.html`) allow users to write valid HTML code alongside
regular markdown text.

```{r}
lines <- c(
  "<table><tr><td>",
  "<pre>",
  "**Hello**,",
  "_world_.",
  "</pre>",
  " </td></tr></table>"
)
md_text(lines)
```

```{r echo=FALSE, results='asis'}
lines <- c(
  "<table><tr><td>",
  "<pre>",
  "**Hello**,",
  "_world_.",
  "</pre>",
  " </td></tr></table>"
)
md_text(lines)
```

### 4.7 Link reference definitions

> A link reference definition consists of a link label, indented up to three
spaces, followed by a colon (`:`), optional whitespace (including up to one line
ending), a link destination, optional whitespace (including up to one line
ending), and an optional link title, which if it is present must be separated
from the link destination by whitespace. No further non-whitespace characters
may occur on the line.
>
> A link reference definition does not correspond to a structural element of a
document. Instead, it defines a label which can be used in reference links and
reference-style images elsewhere in the document. Link reference definitions can
come either before or after the links that use them.

The `md_reference()` function is used to create the link reference definition
_only_, which must then be paired with the same label elsewhere in the document.
We can put the definitions anywhere in the document, so when I type `[CRAN]`
here in this paragraph (like [CRAN]), that text looks for the label created
below to create a hyperlink. The reference definition printed to the document
actually doesn't show up anywhere in the rendered `.html` file.

```{r}
a <- c("cran", "tidy")
b <- c("https://cran.r-project.org/", "https://www.tidyverse.org/")
c <- c("CRAN Home", "Tidyverse Home")
```

```{r}
md_reference(label = a, url = b, title = c)
```

```{r echo=FALSE, results='asis'}
md_reference(label = a, url = b, title = c)
```

### 4.8 Paragraphs

> A sequence of non-blank lines that cannot be interpreted as other kinds of
blocks forms a paragraph. The contents of the paragraph are the result of
parsing the paragraph’s raw content as inlines. The paragraph’s raw content is
formed by concatenating the lines and removing initial and final whitespace.

Paragraphs are one instance are where it's easy for markdown newcomers can have
difficulty writing the plain text in the proper format to produce their desired
outcome. Words separated by a single newline are rendered as a single paragraph
`<p>` tag. Using `md_paragraph()`, we can take a vector of character text and
output _distinct_ paragraphs.

In section 6.12 and 6.13 you can compare the output of `md_paragraph()` with
the other line break functions.

```{r}
sentences <- sample(stringr::sentences, size = 5)
```

```{r}
md_paragraph(sentences)
```

```{r echo=FALSE, results='asis'}
md_paragraph(sentences)
```

### 4.9 Blank lines

> Blank lines between block-level elements are ignored, except for the role they
play in determining whether a list is tight or loose.
> 
> Blank lines at the beginning and end of the document are also ignored.

Blank lines are, well, not super useful... but you can create them with 
`md_blank()`!

```{r}
md_blank()
```

```{r echo=FALSE, results='asis'}
md_blank()
```

### 4.10 Tables (extension)

> GFM enables the `table` extension, where an additional leaf block type is available.
>
> A table is an arrangement of data with rows and columns, consisting of a
single header row, a delimiter row separating the header from the data, and zero
or more data rows.
> 
> Each row consists of cells containing arbitrary text, in which inlines are
parsed, separated by pipes (`|`). A leading and trailing pipe is also
recommended for clarity of reading, and if there’s otherwise parsing ambiguity.
Spaces between pipes and cell content are trimmed. Block-level elements cannot
be inserted in a table.
>
>The delimiter row consists of cells whose only content are hyphens (`-`), and
optionally, a leading or trailing colon (`:`), or both, to indicate left, right,
or center alignment respectively.

The `md_table()` function wraps around `knitr::kable()` to return R objects
(Typically data frames or matrices) as markdown tables. This is useful for
presenting data created using R in a format that's more readable than a
plain-text data frame (or tibble) printout.

```{r message=FALSE, warning=FALSE}
states <- head(state.x77)
print(states)
```

```{r}
options(knitr.kable.NA = "")
md_table(states, align = "c")
```

```{r echo=FALSE, results='asis'}
options(knitr.kable.NA = "")
md_table(states, align = "c")
```

## 5 Container Blocks

> A container block is a block that has other blocks as its contents. There are
two basic kinds of container blocks: block quotes and list items. Lists are
meta-containers for list items.

### 5.1 Block quotes

> A block quote marker consists of 0-3 spaces of initial indent, plus (a) the
character > together with a following space, or (b) a single character > not
followed by a space.

I've been copy-pasting block quotes from the GFM spec in this vignette, but we
can also use `md_quote()` to grammatically create or manipulate strings and
print them as block quotes.

```{r}
read_html("https://w.wiki/A58") %>% 
  html_node("blockquote") %>% 
  html_text(trim = TRUE) %>% 
  str_remove("\\[(.*)\\]") %>% 
  md_quote()
```

```{r echo=FALSE, results='asis'}
read_html("https://w.wiki/A58") %>% 
  html_node("blockquote") %>% 
  html_text(trim = TRUE) %>% 
  str_remove("\\[(.*)\\]") %>% 
  md_quote()
```

```{r}
md_quote(sentences)
```

```{r echo=FALSE, results='asis'}
md_paragraph(md_quote(sentences))
```

### 5.2 List items

> A list marker is a bullet list marker or an ordered list marker.
> 
> A bullet list marker is a -, +, or * character.
>
> An ordered list marker is a sequence of 1–9 arabic digits (`0-9`), followed by
either a `.` character or a `)` character. (The reason for the length limit is
that with 10 digits we start seeing integer overflows in some browsers.)

List items are automatically created in list blocks (section 5.4) via 
`md_order()` and `md_bullet()`.

### 5.3 Task list items (extension) 

> GFM enables the `tasklist` extension, where an additional processing step is
performed on list items.
>
> A task list item is a list item where the first block in it is a paragraph
which begins with a task list item marker and at least one whitespace character
before any other content.
>
> A task list item marker consists of an optional number of spaces, a left
bracket (`[`), either a whitespace character or the letter x in either lowercase
or uppercase, and then a right bracket (`]`).
>
> When rendered, the task list item marker is replaced with a semantic checkbox
element; in an HTML output, this would be an `<input type="checkbox">` element.
>
> If the character between the brackets is a whitespace character, the checkbox
is unchecked. Otherwise, the checkbox is checked.

Markdown extensions like this one are optionally supported on different venues.
On GitHub (in `README.md` files, issues, or comments, etc.) these tasks list
are rendered as check boxes, sometimes even interactable. On other venues, like
this very HTML vignette, these check boxes just render are bullet lists with
weird bracket boxes on each line.

```{r}
md_task(sentences, check = 3:4)
```

```{r echo=FALSE, results='asis'}
md_task(sentences, check = 3:4)
```

### 5.4 Lists

> A list is a sequence of one or more list items of the same type. The list
items may be separated by any number of blank lines.
> 
> Two list items are of the same type if they begin with a list marker of the
same type. Two list > markers are of the same type if (a) they are bullet list
markers using the same character (`-`, `+`, or `*`) or (b) they are ordered list
numbers with the same delimiter (either `.` or `)`).
> 
> A list is an ordered list if its constituent list items begin with ordered
list markers, and a > bullet list if its constituent list items begin with
bullet list markers.
> 
> The start number of an ordered list is determined by the list number of its
initial list item. The numbers of subsequent list items are disregarded.

The `md_order()` and `md_bullet()` functions let users display character vectors
as list blocks. The functions let users adjust the which list item markets are
used, but all the options are rendered as the same `<ul>` or `<ol>` tags in the
rendered `.html` file.

```{r}
states <- sample(state.name, size = 3)
```

```{r}
md_bullet(states)
```

```{r echo=FALSE, results='asis'}
md_bullet(states)
```

```{r}
md_bullet(states, marker = "+")
```

```{r echo=FALSE, results='asis'}
md_bullet(states, marker = "+")
```

```{r}
md_order(states)
```

```{r echo=FALSE, results='asis'}
md_order(states)
```

```{r}
md_order(states, seq = FALSE, marker = ")")
```

```{r echo=FALSE, results='asis'}
md_order(states, seq = FALSE, marker = ")")
```

## 6 Inlines

> Inlines are parsed sequentially from the beginning of the character stream to
the end (left to right, in left-to-right languages). Thus, for example, in

### 6.1 Backslash escapes

> Any ASCII punctuation character may be backslash-escaped... Escaped characters
are treated as regular characters and do not have their usual Markdown meanings:

If you want to display markdown syntax without rendering as the corresponding
HTML tags, `md_escape()` can prepend a backslash (actually, two) before every
ASCII punctuation character.

```{r}
inline <- md_bold("example")
md_escape(inline)
```

```{r echo=FALSE, results='asis'}
md_escape(inline)
```

### 6.3 - 6.8 Simple inlines

Here we are going to demonstrate how various simple inline block functions are
used to create text _within_ a container block.

```{r}
inlines <- c(
  md_code("6.3 Code spans"),
  md_italic("6.4 Emphasis"),
  md_bold("6.4 Strong emphasis"),
  md_strike("6.5 Strikethrough (extension)"),
  md_link("6.6 Links", "https://example.com/"),
  md_autolink("https://example.com/")
)
```

```{r}
md_bullet(inlines, marker = "-")
```

```{r echo=FALSE, results='asis'}
md_bullet(inlines, marker = "-")
```

### 6.7 Images 

> Syntax for images is like the syntax for links, with one difference. Instead
of link text, we have an image description. The rules for this are the same as
for link text, except that (a) an image description starts with `![` rather than
`[`, and (b) an image description may contain links. An image description has
inline elements as its contents. When an image is rendered to HTML, this is
standardly used as the image’s `alt` attribute.

```{r}
md_image("https://www.r-project.org/Rlogo.png", alt = "R logo")
```

```{r echo=FALSE, results='asis'}
tmp <- tempfile(fileext = ".png")
try <- tryCatch(
  expr = download.file("https://www.r-project.org/Rlogo.png", tmp),
  error = function(e) return(NULL)
)
if (!is.null(try)) {
  md_image(tmp, alt = "R logo")
}
```

### 6.11 Disallowed Raw HTML (extension)

> GFM enables the tagfilter extension, where the following HTML tags will be
filtered when rendering HTML output:
> 
> * `<title>`
> * `<textarea>`
> * `<style>`
> * `<xmp>`
> * `<iframe>`
> * `<noembed>`
> * `<noframes>`
> * `<script>`
> * `<plaintext>`
>
> Filtering is done by replacing the leading < with the entity &lt;. These tags
are chosen in particular as they change how HTML is interpreted in a way unique
to them (i.e. nested HTML is interpreted differently), and this is usually
undesireable in the context of other rendered Markdown content.

```{r}
lines <- c(
  "<blockquote>",
  "  <xmp> is disallowed.  <XMP> is also disallowed.",
  "</blockquote>"
)
md_disallow(lines)
```

```{r echo=FALSE, results='asis'}
lines <- c(
  "<blockquote>",
  "  <xmp> is disallowed.  <XMP> is also disallowed.",
  "</blockquote>"
)
md_disallow(lines)
```

All other HTML tags are left untouched.

### 6.12 - 6.13 Line breaks

The difference between _hard_ line breaks, _soft_ line breaks, paragraphs,
and regular text can be confusing. New markdown users often struggle with
knowing exactly how many invisible newlines or spaces are needed to present
each line in the way they want. The line break functions below make it easier.

> A line break (not in a code span or HTML tag) that is preceded by two or more
spaces and does not occur at the end of a block is parsed as a hard line break
(rendered in HTML as a `<br />` tag):

```{r}
# 6.12 Hard line breaks
md_hardline(sentences)
```

```{r echo=FALSE, results='asis'}
md_hardline(sentences)
```

***

> A regular line break (not in a code span or HTML tag) that is not preceded by
two or more spaces or a backslash is parsed as a softbreak. (A softbreak may be
rendered in HTML either as a line ending or as a space. The result will be the
same in browsers. In the examples here, a line ending will be used.)

```{r}
# 6.13 Soft line breaks
md_softline(sentences)
```

```{r echo=FALSE, results='asis'}
md_softline(sentences)
```

***

You can see how both hard line and soft line breaks are different from the
paragraphs as created using `md_paragraph()`

```{r}
md_paragraph(sentences)
```

```{r echo=FALSE, results='asis'}
md_paragraph(sentences)
```

### 6.14 Textual content

> Any characters not given an interpretation by the above rules will be parsed
as plain textual content.

The `md_text()` function is a catch-all wrapper around `glue::as_glue()` with
simply converts any character string into a `glue` object, which when used in
an `.Rmd.` document alongside the `results='asis'` allows users to simply print
text to the body of the document.

```{r}
# 6.14 Textual content
md_text(sentences)
```

```{r echo=FALSE, results='asis'}
md_text(sentences)
```

***

Compare this to the regular handling of character strings.

```{r}
print(sentences)
```

```{r echo=FALSE, results='asis'}
print(sentences)
```

***

The `md_text()` function creates strings with the `glue` S3 class. This class
works in the same way as the `cat()` functions.

```{r}
cat(sentences)
```

```{r echo=FALSE, results='asis'}
cat(sentences)
```
