---
title: "Markdown Literal Programming"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Printing Markdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: console
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r libraries, warning=FALSE, message=FALSE, error=FALSE}
library(gluedown)
library(stringr)
library(rvest)
library(glue)
```

In literate programming, the typical paradigm of source code is reversed;
instead of a wall of code with the occasional comment, the user writes _human_
readable text (like this paragraph) with source code interspersed. In the R
language, this is primarily done with the [`rmarkdown`][rmd] package, which
takes a plaint text _R_ markdown file (`.Rmd`) containing code "chunks" and
executes that code when converting to a regular markdown file (`.md`) and then
possibly some other format (`.html`, `.pdf`, etc).

[rmd]: https://rmarkdown.rstudio.com/

Markdown is a lightweight plain-text language used to format text. Let's look at
the original description of markdown from John Gruber's website, the creator of
the markdown standard. Using the `rvest` package, we can programmatically scrape
Gruber's blog, extract HTML paragraph tags, and convert those tags to character
vectors.

```{r markdown_desc, eval=FALSE}
markdown_blog <- 
  read_html("https://daringfireball.net/projects/markdown/") %>% 
  html_elements("p") %>% 
  html_text()
```

```{r echo=FALSE}
x <- tryCatch(
  expr = read_html("https://daringfireball.net/projects/markdown/"),
  error = function(e) NULL
)
if (!is.null(x)) {
 markdown_blog <- html_text(html_elements(x, "p"))
} else {
  markdown_blog <- LETTERS
}
```

Gruber first explains _what_ exactly his markdown language is.

```{r quote_what, results='asis'}
md_quote(markdown_blog[4])
```

He continues by outlining _why_ markdown was created, his rationale for it's
format, and some inspiration for it's syntax.

```{r quote_why, results='asis'}
md_quote(markdown_blog[6])
```

This entire vignette was written in markdown and converted to HTML using 
[pandoc][pandoc]. However, as you may have noticed, we haven't exactly been
conforming to this original desire for markdown to be readable as is. We didn't
copy the text from his blog and past it as text into this vignette. This is
where the `gluedown` package comes in.

[pandoc]: https://pandoc.org/

The `gluedown` package helps ease the transition between the incredibly powerful
vector support in R and the readability of markdown. Since this vignette was
written in _R_ Markdown (`.Rmd`), we are able to (1) use the power of packages
like `rvest` to collect, process, and/or analyze some kind of data and then (2)
transition that result to the human readable markdown format.

When writing this vignette, _three_ kinds of files are used.

1. The `.Rmd` file containing source code is a programming environment
2. The `.md` file created by `rmarkdown` is a human-readable plain text version
of that input code also containing the output text.
3. The `.html` format of this vignette created by `pandoc` is the final
presentation format.

In the rest of this vignette, we will see some of the various use cases for
`gluedown`. We will see how easy it is to transition between R vectors and
readable results in markdown/HTML.

## Vector Lists

Printing vectors as markdown lists was the initial inspiration for the package.
In R, atomic vectors the fundamental object type that composes more complex
objects like lists and dataframes. The `state.name` vector built into base R
is a character vector of all 50 state names.

```{r state.name}
str(state.name, vec.len = 3)
```

If we as a user want to use those state names as _text_ in our markdown document
we can use the `cat()` function and tell `rmarkdown` to print the results of
that function "as is" (rather than as code output).

```{r cat_plain, results='asis'}
cat(state.name[1:3])
```

That output obviously isn't very appealing. We could tweak our use of `cat()` a
little to separate them on new lines.

```{r cat_newline, results='asis'}
cat(state.name[1:3], sep = "\n\n")
```

This is more readable, but with some more work, we can use `cat()` to print an
ordered list.

```{r cat_order, results='asis'}
cat(paste0(1:3, ". ", state.name[1:3]), sep = "\n")
```

This workflow gets tiresome, although it's made slightly more simple with
the fantastic [`glue`][glue] package from Jim Hester.

```{r glue_order, results='asis'}
glue("{1:3}. {state.name[1:3]}")
```

This is the technique used in this package. Vector inputs are passed to
`glue::glue()` and the appropriate markdown syntax is implemented.

[glue]: https://github.com/tidyverse/glue

The `md_order()` function simplifies the `glue::glue()` workflow and allows
users to more easily customize the appearance of the list in _markdown_ format.

```{r order_list_raw}
# markdown only cares about the first number
md_order(state.name[1:3], seq = FALSE)
# markdown ignored padding and allows for use of parentheses
md_order(state.name[1:10], seq = TRUE, pad = TRUE, marker = ")")
```

Although, as we can see below, all these different options are rendered as the
same kind of HTML `<ol>` fragment.

```{r order_list_pretty, results='asis'}
md_order(state.name[1:3], seq = FALSE)
md_order(state.name[1:10], seq = TRUE, pad = TRUE, marker = ")")
```

This ordered list is a markdown container _block_. As described in the GitHub
Flavored Markdown specification:

> We can think of a document as a sequence of blocks—structural elements like
paragraphs, block quotations, lists, headings, rules, and code blocks. Some
blocks (like block quotes and list items) contain other blocks; others (like
headings and paragraphs) contain inline content—text, links, emphasized text,
images, code spans, and so on.

We can nest `md_*()` functions to create inline content within a code block.
Let's use some inline functions to create a new vector names `inline` with
five states, each formatted in another syntax. We'll take a look at what that
vector _really_ looks like with a simple `print()`.

```{r inline_vector}
inlines <- c(
  md_bold(state.name[4]),
  md_code(state.name[5]),
  md_link(state.name[6], "https://Colorado.gov"),
  md_italic(state.name[7]),
  md_strike(state.name[8])
)

str(inlines, vec.len = 3)
```

Using `md_bullet()` we will print that vector as a bullet point list container
block and each list item will be rendered as a separate inline.

```{r bullet_list, results='asis'}
md_bullet(inlines)
```

These functions demonstrate how `gluedown` can be used to transition between
R vectors, simply formatted markdown text, and beautifully formatted HTML text.

Aside from container blocks and inlines, there is a third type of markdown
content. The leaf blocks cannot contain inline content. The thematic break
is an example of a leaf block.

```{r theme_break, results='asis'}
md_rule(char = "*", n = 80)
```

Code blocks are another type of leaf block. The code we've been writing so far
is contained within `rmarkdown` _chunks_, which execute the code within. By
default, those code chunks are then displayed as regular code blocks in the
intermediary `.md` file. Sometimes we might want to use code blocks to display
other types of text. Perhaps we want to show the content of a function. The
`md_fence()` function creates a _new_ code fence from the lines created by
`deparse()`.

```{r fence_block, results='asis'}
lines <- deparse(md_bullet)
md_fence(lines)
```

Or perhaps we want to display some code from another language that isn't
supposed to be executed

```{r tilde_block}
command <- "sudo apt install r-base-dev"
md_fence(paste("$", command), char = "~", info = "bash")
```

## Pipes

The package has been designed to fit well in a traditional R workflow so users
can seamlessly create content with their code and display that content with
`gluedown`. In that spirit, all functions are designed to fit within the
tidyverse ecosystem by working with [pipes][pipe]. Pipes allow users to pass
the results of one function into the beginning of the next. By ending this
"pipeline" with `md_quote()`, we chain together five coding steps:

1. Read the HTML text of a Wikipedia page
1. Extract the first `<blockquote>` tag
1. Convert that tag to a character vector
1. Remove Wikipedia's bracketed note
1. Print that vector as a markdown block quote

```{r blockquote, results='asis', eval=FALSE}
read_html("https://w.wiki/A58") %>% # 1
  html_element("blockquote") %>%    # 2
  html_text(trim = TRUE) %>%        # 3
  str_remove("\\[(.*)\\]") %>%      # 4
  md_quote()                        # 5
```

```{r results='asis', echo=FALSE}
w <- "https://en.wikipedia.org/wiki/Preamble_to_the_United_States_Constitution"
x <- tryCatch(
  expr = read_html(w),
  error = function(e) NULL
)
if (!is.null(x)) {
  x %>%
    html_element("blockquote") %>%
    html_text(trim = TRUE) %>%
    str_remove("\\[(.*)\\]") %>%
    md_quote() 
}  
```

[pipe]: https://magrittr.tidyverse.org/reference/pipe.html

## Extensions

The package primarily uses [GitHub Flavored Markdown][gfm] (GFM), a
site-specific version of the [CommonMark specification][cm], an unambiguous
implementation of John Gruber's [original Markdown][df]. With this flavor, some
useful extensions like [task lists][task] are supported on GitHub. Elsewhere,
like this HTML vignette, a task list will just render as a bullet list. You can
learn more about how GFM us implemented in this package's other vignette.

```{r ex_task, results='asis'}
legislation <- c("Houses passes", "Senate concurs", "President signs")
md_task(legislation, check = 1:2)
```

[task]: https://help.github.com/en/articles/about-task-lists
[gfm]: https://github.github.com/gfm/
[cm]: https://spec.commonmark.org/
[df]: https://daringfireball.net/projects/markdown/

Markdown tables are another extremely useful extension. The `md_table()`
functions wraps around the much more powerful `knitr::kable()` function, which
allows data frames to be printed in a number of alternative formats. Printing
data frames is a very typical use case for documenting the process of data
science. With small summary tables like the one below, a markdown table is much
more readable than the plain text tibble or data frame printed by default.

```{r print_mass}
print(head(state.x77))
```

```{r table_mass, results='asis'}
md_table(head(state.x77), digits = 2)
```

## Inlines

You can also use `gluedown` to format R [inline code results][inline]. First,
use R to calculate a result.

```{r ex_inline}
rand <- sample(state.name, 1)
# `r md_bold(rand)`
var <- sample(colnames(state.x77), 1)
# `r md_code(var)`
```

Then, you can easily print that result in the middle of regular text with
markdown formatting. In this case, our randomly selected state is... 
`r md_bold(rand)` and the `r md_code(var)` variable was randomly selected from
the `state.x77` dataframe. Calculating results and using those
calculations in the body of a text document increases reproducibility. 

In a [meta-study of psychology journals][pmc], researchers found that "around
15% of the articles contained at least one statistical conclusion that proved,
upon recalculation, to be incorrect." These errors can be mitigated by using
inline printing of results like we did above. With the `gluedown` package,
programmers can **emphasize** those results without worry.

[pmc]: https://www.ncbi.nlm.nih.gov/pmc/articles/PMC3174372/
